---
title: "Checkerboard - regression"
author: "Diana Proctor"
date: "2025-07-30"
output: html_document
---


---
last updated: `r format(Sys.Date(), format="%B %d %Y")`

***


chatGPT was used to help code this up

note: this is to analyze end-point data, not a time course!!!! here, we look at the 24h readings

```{r}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=8, 
               echo=TRUE, warning=FALSE, message=FALSE, error = FALSE)
```

Load R packages
```{r}
# Set seed
set.seed(78979)

# Define required packages
packages <- c(
  # Core tidyverse packages
  "tidyverse",
  
  # Phylogenetics and microbiome analysis
  "phyloseq",
  "microbiome",
  "phangorn",
  "msa",
  
  # Statistical modeling and analysis
  "lmerTest",
  "rstatix",
  "broom",
  "mediation",
  "effects",
  "broom.mixed",
  "genefilter",
  "vegan",
  "ALDEx2",
  
  # Visualization enhancements
  "ggpubr",
  "ggprism",
  "gridExtra",
  "ggbeeswarm",
  "ggrepel",
  "cowplot",
  "ggtext",
  "reshape2",
  "pheatmap",
  "ggfortify",
  
  # Data import
  "readxl"
)

# Install missing CRAN packages
cran_packages <- setdiff(packages, rownames(installed.packages()))
cran_packages <- cran_packages[!cran_packages %in% c("phyloseq", "microbiome", "phangorn", "msa", "genefilter", "vegan", "ALDEx2")]
if(length(cran_packages) > 0) install.packages(cran_packages)

# Install missing Bioconductor packages
bioc_packages <- c("phyloseq", "microbiome", "phangorn", "msa", "genefilter", "vegan", "ALDEx2")
bioc_to_install <- setdiff(bioc_packages, rownames(installed.packages()))
if(length(bioc_to_install) > 0) BiocManager::install(bioc_to_install, update = FALSE)

# Load packages
n <- length(packages) - sum(sapply(packages, require, character.only = TRUE))

# Print loading status
if (n == 0) {
  print("All necessary R packages loaded properly")
} else {
  print(paste0(n, " R packages did not load properly"))
}

#set graphs to the graphpad prism theme
theme_set(theme_bw())

```

read in the data and merge it together

specifically, we read in the replicates 
```{r}
process_checkerboard_data <- function(file_path, sheet = 1) {
  # Read the Excel sheet
  raw_data <- read_excel(file_path, sheet = sheet)
  
  # Subset the data from row 26 onward
  data_subset <- as.data.frame(raw_data[26:nrow(raw_data), ])
  
  # Remove the first column
  data_subset <- data_subset[, -1]
  
  # Remove the first row
  cleaned_data <- data_subset[-1, ]
  cleaned_data$file = file_path
  return(cleaned_data)
}

file1 <- "~/Desktop/Staff/Sanduni/Checkerboard assays_Van_Flu_old RPMI/Vanco_Fluco_Pilot Experiment_19_05_2024.xlsx"
rep1 <- process_checkerboard_data(file1)


file2 <- "~/Desktop/Staff/Sanduni/Checkerboard assays_Van_Flu_old RPMI/Vanco_Fluco_Replicate 1_24_05_2024.xlsx"
rep2 <- process_checkerboard_data(file2)


file3 <- "~/Desktop/Staff/Sanduni/Checkerboard assays_Van_Flu_old RPMI/Vanco_Fluco_Replicate 2_ Reading 1_24_05_2025.xlsx"
rep3 <- process_checkerboard_data(file3)

file4 <- "~/Desktop/Staff/Sanduni/Checkerboard assays_Van_Flu_old RPMI/Vanco_Fluco_Replicate2 _Reading 2_24_05_2025.xlsx"
rep4 <- process_checkerboard_data(file4)

#make a dataset of all the data
od_df = data.frame(rbind(rep2, rep3, rep4)) #need to add rep1 

#melt the data
colnames(od_df) = c("Row", 1:12, "wv", "file")
# Pivot longer
od_long <- od_df %>%
  pivot_longer(
    cols = `1`:`12`,         # melt columns 1 to 12
    names_to = "Col",     # new column with the original column names
    values_to = "OD"         # new column with OD values
  )


```

read in the metadata

```{r, chunk from sanduni}
# Load metadata
file="~/Desktop/Staff/Sanduni/Checkerboard assays_Van_Flu_old RPMI/Checkerboard_Van_Flu_23_06_25 @24hr.xlsx"
meta_df <- read_excel(file, sheet =2)

# Rename row identifiers
colnames(meta_df)[1] <- "Row"
```

merge the data and clean it up
```{r, sanduni code}
meta_long <- meta_df %>%
  pivot_longer(-Row, names_to = "Col", values_to = "Condition")

# Combine OD + metadata
data <- left_join(od_long, meta_long, by = c("Row", "Col")) %>%
  filter(!is.na(OD), !grepl("EMPTY", OD, ignore.case = TRUE), !is.na(Condition)) %>%
  mutate(OD = as.numeric(OD))

# Extract Van and Flu concentrations
data2 <- data %>%
  mutate(
    FluRaw = str_extract(Condition, "Flu\\s*[0-9.]+"),
    FluConc = as.numeric(str_extract(FluRaw, "[0-9.]+")),
    VanRaw = str_extract(Condition, "Van\\s*[0-9.]+"),
    VanConc = as.numeric(str_extract(VanRaw, "[0-9.]+"))
  )

```

merge OD with pH data
```{r}

#read in the pH data
phdata = read_excel("~/Desktop/ABX_Project/pH and Matrices.xlsx")
phdata_long <- phdata %>%
  dplyr::rename(Van = `Van/Flu`) %>%
  pivot_longer(
    cols = -Van,
    names_to = "Flu",
    values_to = "pH"
  ) %>%
  mutate(Flu = as.numeric(Flu))  

#merge pH data with the OD data
mydata = merge(phdata_long, data2  ) 
#View(mydata)


mydata = subset(mydata, Condition !="EMPTY")
#look at predictors
table(mydata$pH) #pH varies between 5 and 6; it's a qualitative assessment not numeric

#how correlated are they? 
cor(mydata$Van, as.numeric(as.character(mydata$pH))) #very 
```

let's look at the data
```{r}
p = ggplot(mydata, aes(sqrt(pH), sqrt(Van))) + geom_point()
p= ggplot(mydata, aes(as.factor(Van), OD)) + geom_boxplot() + facet_wrap(~Flu)
p= ggplot(mydata, aes(as.factor(pH), OD)) + geom_boxplot() + facet_wrap(~Flu)
```

normalize within plates - modified the code sanduni wrote
we will also just plot the raw data
```{r}
# Get control OD values per file and FluConc
control_vals <- data2 %>%
  filter(
    str_detect(Condition, regex("Control", ignore_case = TRUE)) &
    str_detect(Condition, regex("Van 0", ignore_case = TRUE))
  ) %>%
  group_by(file, FluConc) %>%
  summarise(Control_OD = mean(OD, na.rm = TRUE), .groups = "drop")

#Join back to full dataset
data3 <- left_join(data2, control_vals, by = c("file", "FluConc")) %>%
  mutate(GrowthPercent = (OD / Control_OD) * 100)

# create heatmap matrix per file (list of matrices)
heatmap_list <- data3 %>%
  group_by(file, VanConc, FluConc) %>%
  summarise(Growth = mean(GrowthPercent, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(FluConc), FluConc != 0) %>%   # remove NA and zero fluconazole columns
  group_split(file) %>%
  setNames(unique(data3$file)) %>%
  lapply(function(df) {
    # Pivot so rows = VanConc, columns = FluConc, values = Growth
    mat_data <- df %>%
      pivot_wider(names_from = FluConc, values_from = Growth) %>%
      arrange(VanConc)         # sort rows by VanConc ascending
    
    # Remove rows with NA VanConc (if any)
    mat_data <- mat_data %>% filter(!is.na(VanConc))
    
    # Remove columns that are all NA (sometimes happens if a FluConc had no data)
    na_cols <- sapply(mat_data, function(col) all(is.na(col)))
    mat_data <- mat_data[, !na_cols]
    
    # Extract matrix without the VanConc column
    mat <- as.matrix(mat_data[, !(names(mat_data) %in% c("VanConc", "file"))])

    
    # Ensure matrix is numeric (important for plotting)
    mode(mat) <- "numeric"
    
    # Assign VanConc values as row names
    rownames(mat) <- as.character(mat_data$VanConc)
    
    return(mat)
  })

########plot the heatmaps for each file
color_breaks <- seq(0, 300, length.out = 100)  # fixed color scale from 0 to 300

heatmap_plots <- list()

for (file_name in names(heatmap_list)) {
  file_base <- tools::file_path_sans_ext(basename(file_name))
  
  heatmap_obj <- pheatmap(
    heatmap_list[[file_name]],
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    main = file_base,
    silent = TRUE,
    breaks = color_breaks
  )
  
  heatmap_plots[[file_base]] <- heatmap_obj[[4]]
}
combined_plot <- cowplot::plot_grid(plotlist = heatmap_plots, ncol = 2)  # or adjust ncol
combined_plot

```


let's get the average of the three replicates
```{r} 
# Assuming heatmap_list contains matrices of identical dimensions

# Convert list to 3D array: [row, col, replicate]
heatmap_array <- simplify2array(heatmap_list)

# Compute mean across the third dimension (replicates)
average_matrix <- apply(heatmap_array, c(1, 2), mean, na.rm = TRUE)

# View or return the result
print(average_matrix)
avg_plot = pheatmap(average_matrix,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    main = file_base,
    silent = TRUE,
    breaks = color_breaks)
avg_plot
```
 
we can see that vancomycin explained 12% of the variation while pH explains 15%. in the combined model, we only account for 22% of the variation, which is less than 12+15. This confirms that the effect of ph and the effect of vancomycin overlap to some degree and account for a shared proportion of the explained variance in the combined model

```{r}
#merge pH with data3
colnames(phdata_long) = c("VanConc", "FluConc", "pH")
data4 = merge(data3, phdata_long)

# Fit the models (note we will use hierarchical regression)
mod_van <- lm(GrowthPercent ~ VanConc+FluConc, data = data4) #vancomycin alone
mod_ph <- lm(GrowthPercent ~ pH+FluConc, data = data4) #pH alone
mod_van_ph <- lm(GrowthPercent ~ VanConc + pH + FluConc, data = data4) #both 

# Extract R-squared values
R2_van <- summary(mod_van)$r.squared
R2_ph <- summary(mod_ph)$r.squared
R2_van_ph <- summary(mod_van_ph)$r.squared

# Calculate unique and shared variance components
unique_van <- R2_van_ph - R2_ph          # variance Van explains beyond pH
unique_ph  <- R2_van_ph - R2_van         # variance pH explains beyond Van
shared     <- R2_van + R2_ph - R2_van_ph # overlap in variance explained

# Print results
cat("R² Van alone:", 100*(round(R2_van, 4)), "\n")
cat("R² pH alone:", 100*(round(R2_ph, 4)), "\n")
cat("R² Van + pH:", 100*(round(R2_van_ph, 4)), "\n\n")

cat("Unique variance explained by Van:", 100*(round(R2_van, 4)), "\n")
cat("Unique variance explained by pH:", 100*(round(R2_ph, 4)), "\n")
cat("Shared variance explained by both:", 100*(round(R2_van_ph, 4)), "\n")
```

let's test for interactions between van and pH while holding fluconazole constant

- when we do this, vancomycin is not significant
- the interaction between van and ph is marginally significant with a positive effect size, indicating a positive effect on growth and that the effect of vancomycin depends on ph or vice versa
- the effect of flu is significant and negative, as expected
- when we look at just pH 6, higher vancomycin concentrations is associated with higher growth
```{r}
mod_interact <- lm(GrowthPercent ~ VanConc * FluConc * pH, data = data4)
summary(mod_interact)


#try to look at the of vancomycin and ph across fluconazole
ggplot(data4, aes(x = VanConc, y = sqrt(GrowthPercent), color = as.factor(pH))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + facet_wrap(~FluConc)

#try to look at how vancomycin and fluconazole interact across ph gradient
ggplot(data4, aes(x = pH, y = GrowthPercent, color = as.factor(VanConc))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + facet_wrap(~FluConc) #when we look at just pH 6, higher vancomycin concentrations is associated with higher growth
```


let's test whether the vancomycin-fluconazole interaction is significant when we hold pH consta nt

```{r}
mod_growth <- lm(GrowthPercent ~ VanConc * FluConc + pH, data = data4)
summary(mod_growth)
```

chatGPT suggested using this to plot interactions
```{r}
plot(allEffects(mod_growth))
```

let's standardize the predictors
```{r}
data4$Van_std <- scale(data4$VanConc)
data4$pH_std <- scale(data4$pH)
data4$Flu_std <- scale(data4$FluConc)
mod_interact <- lm(GrowthPercent ~ Van_std * pH_std * Flu_std, data = data4)
summary(mod_interact)
autoplot(mod_interact)
```

```{r}
p = ggplot(data4, aes(as.factor(pH), GrowthPercent)) + geom_boxplot() + geom_point()
p
p = ggplot(data4, aes(as.factor(pH), GrowthPercent)) + geom_boxplot() + geom_point() + facet_wrap(~FluConc)
p
p = ggplot(data4, aes(as.factor(VanConc), GrowthPercent, shape=as.factor(pH), color=as.factor(FluConc))) + geom_point(size=3) 
p
```


### try to permute vancomycin concentrations

While holding pH and Fluconazole concentration constant in the model, the observed effect of Vancomycin concentration on growth percentage was substantially greater than expected under the null hypothesis. In a permutation test with 1000 random shuffles of Vancomycin concentration, only 4 permutations produced p-values smaller than the observed p-value, indicating that the likelihood of observing such a strong effect by chance is very low (empirical p-value = 0.004), much lower than the typical 5% threshold used to delcare significance. This provides strong evidence that Vancomycin concentration has a significant influence on growth, independent of the effects of pH and Fluconazole, and is unlikely to be a false positive.



```{r}
set.seed(123)
n_perms <- 1000

# Vectors/lists to store permutation results
vancoefs <- numeric(n_perms)  # VanConc coef from each permuted model
van_p_vals <- numeric(n_perms) # p values from permuted model 
perm_resid_vs_fitted <- vector("list", n_perms)

for (i in 1:n_perms) {
  data4$VanConc_perm <- sample(data4$VanConc)  # shuffle VanConc

  # Fit permuted model
  perm_mod <- lm(GrowthPercent ~ VanConc_perm + pH + FluConc, data = data4)
  
  # Store VanConc coefficient (permuted)
  vancoefs[i] <- coef(perm_mod)["VanConc_perm"]
  van_p_vals[i] <- summary(perm_mod)$coefficients["VanConc_perm", "Pr(>|t|)"]
  # Store Residuals vs Fitted data for diagnostics
  perm_resid_vs_fitted[[i]] <- data.frame(
    Fitted = fitted(perm_mod),
    Residuals = resid(perm_mod),
    Permutation = i
  )
}

# Fit original model (unpermuted)
obs_mod <- lm(GrowthPercent ~ VanConc + pH + FluConc, data = data4)
obs_coef <- coef(obs_mod)["VanConc"]
obs_p <- summary(obs_mod)$coefficients["VanConc", "Pr(>|t|)"]

# Plot histogram of permuted VanConc coefficients with observed value line
p1 = ggplot(data.frame(vancoefs), aes(x = vancoefs)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  geom_vline(xintercept = obs_coef, color = "red", size = 1.2) +
  labs(title = "Permutation distribution of VanConc coefficients",
       x = "VanConc coefficient",
       y = "Frequency") +
  theme_minimal()

#plot the p values
p2 = ggplot(data.frame(van_p_vals), aes(x = van_p_vals)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  geom_rect(aes(xmin = 0, xmax = 0.05, ymin = 0, ymax = Inf),
            fill = "pink", alpha = 0.05, inherit.aes = FALSE) +
  labs(title = "Permutation distribution of VanConc p-values",
       x = "VanConc p-value",
       y = "Frequency") +
  theme_minimal() +  geom_vline(xintercept = obs_p, color = "red", size = 1.2) 


plot_grid(p1, p2, ncol=2)
###########how many permuted p values were < 0.05
num_less_than_obs <- sum(van_p_vals < obs_p)
total_perms <- length(van_p_vals)

cat("Out of", total_perms, "permutations,", num_less_than_obs, 
    "had p-values less than the observed p-value of", round(obs_p, 5), ".\n")

if (num_less_than_obs == 0) {
  cat("None of the permutations had a p-value smaller than the observed p-value — strong evidence against the null hypothesis.\n")
} else {
  cat("This means about", round(100 * num_less_than_obs / total_perms, 2), 
      "% of permutations had smaller p-values than the observed one,\n")
}

```


let's try to remove the effect of pH and then run our regression model. susan said to do this, but i should make sure i'm doing it right by checking with julia. this suggests that the effect of vancomcyin after controlling for pH is negative

```{r}
mod_ph <- lm(pH ~ VanConc, data = data4)
data4$pH_resid <- resid(mod_ph)

# Now pH_resid is the part of pH not explained by VanConc
mod <- lm(GrowthPercent ~ VanConc + pH_resid, data = data4)
summary(mod)

```
